% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\documentclass[
  11pt,
]{article}
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage[]{libertinus}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{setspace}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{longtable,booktabs,array}
\newcounter{none} % for unnumbered tables
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
% Make links footnotes instead of hotlinks:
\DeclareRobustCommand{\href}[2]{#2\footnote{\url{#1}}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\author{}
\date{}

\begin{document}

\setstretch{1.2}
\section{Abstract}\label{abstract}

The traditional chronology of the British Industrial Revolution credits
fossil fuels---specifically coal extraction and the steam engine---as
the primary catalysts that liberated human economics from the
constraints of nature. This paper refines the institutional
``steam-first'' timeline by demonstrating that a measurable structural
break in Britain's macroeconomic trajectory originated during the
earlier era of water infrastructure, preceding and establishing the
preconditions for widespread fossil adoption. By applying Natural
Language Processing (NLP) to the historical Google Books
\texttt{eng\_gb\_2019} corpus, we identify the exact structural shift
wherein Britain's printed vocabulary transitioned from viewing water as
a natural hazard to commodifying it as an engineered asset (the
``Hydro-Social Shift'').

Our analysis leverages the 1761 opening of the Bridgewater Canal as a
definitive exogenous infrastructure shock. Deploying 1761 as the
treatment threshold within a robust Difference-in-Differences (DiD)
framework using Maddison Project historical GDP data, we find that
Britain's post-shock economic trajectory diverged by an additional
\(1,251 GDP per capita relative to continental European controls (\)p =
0.042\$, HAC). We validate the mechanism behind this shock using NLP: we
trace the \texttt{eng\_gb\_2019} corpus and identify 1766 as the exact
structural crossover wherein Britain's printed vocabulary permanently
transitioned from naturalistic to engineered water terminology.
Crucially, 47\% of Britain's ultimate industrial divergence was
established during this pre-steam ``Water Era,'' decades before steam
engines achieved commercial dominance. Placebo falsification tournaments
across alternative industrial and agricultural vocabularies yield noisy
or null effects, demonstrating that only the water infrastructure shock
produces a clean event study aligned with GDP divergence. The data
positions the canal era not as the sole engine of industrial modernity,
but as a necessary precondition---an era of ``geographical symbiosis''
that created the integrated markets, accumulated capital, and systemic
demand which the subsequent fossil transition would amplify.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\section{1. Introduction}\label{introduction}

The origins of the British Industrial Revolution have been heavily
debated, yet prominent perspectives often gravitate toward a singular
technological rupture: the adoption of coal and the maturation of the
steam engine. According to this narrative---deeply rooted in the
\emph{Unbound Prometheus} paradigm formalized by David S. Landes (1969)
and enduring across both classical Marxist and neoclassical
historiography---macroeconomic modernity accelerated significantly when
economies broke free from the natural limitations of the organic economy
by extracting subterranean fossil fuels. Within this framework,
industrialization is frequently characterized philosophically and
economically as humanity's uncoupling from, and mastery over, the
natural environment.

This perspective is entrenched in the historiography of economic
development. E.A. Wrigley famously conceptualized the Industrial
Revolution as the necessary transition from an ``organic
economy''---limited by the photosynthetic capture of solar energy via
wood and wind---to a ``mineral-based energy economy'' built on coal
(Wrigley 2010). In his view, sustained exponential growth was physically
impossible within the confines of organic flows. Similarly, Robert
Allen's robust ``High Wage Economy'' thesis posits that Britain's unique
matrix of cheap coal and high labor costs structurally induced the
invention of the steam engine, treating geological luck as the prime
engine of British divergence (Allen 2009).

However, this energy-centric perspective operates alongside equally
foundational institutional and cultural theories. Deirdre McCloskey
argues that an epistemological shift in rhetoric and culture---how
society spoke about commerce and innovation---was a crucial prerequisite
for industrialization (McCloskey 2010). Similarly, Joel Mokyr theorizes
that exponential growth was unlocked by the systematic accumulation and
application of ``useful knowledge'' (Mokyr 2009). Yet, unifying these
cultural shifts with the material and energetic transitions of the
period remains an ongoing methodological challenge.

Furthermore, the material realities of early manufacturing complicate
the coal-centric timeline. The sprawling networks of navigable canals,
aqueducts, and mechanized water wheels suggest that the foundation of
the British economic trajectory was laid not by subverting the landscape
via fossil extraction, but by actively partnering with it. As Andreas
Malm argues in \emph{Fossil Capital}, the eventual transition from water
to steam power in the mid-19th century was not driven by thermodynamic
superiority or absolute scarcity of water, but by the socio-spatial
demands of capital. Steam engines allowed factories to be relocated to
urban centers where labor could be disciplined, whereas water power
required factories to adapt to remote riverine ecologies (Malm 2016). If
the energetic foundation of modernity was actually established in the
organic era, the timeline of industrialization is fundamentally late.

This paper examines whether the linguistic commodification of water
preceded the semantic integration of fossil fuels, and whether this
temporal precedence is systematically associated with the initial phase
of Britain's macroeconomic takeoff. By doing so, it serves as an
empirical bridge between McCloskey's rhetoric, Mokyr's useful knowledge,
and the tangible energetic realities of early infrastructure. We propose
that water infrastructure functioned as a necessary precondition for
industrial modernity---creating the integrated markets, capital
accumulation, and systemic demand that the subsequent fossil transition
would amplify---rather than representing an alternative to it.

By deploying a dual-pronged computational methodology---merging
unsupervised natural language processing (NLP) of centuries of
historical print culture with Difference-in-Differences (DiD)
econometric modeling of historical GDP---we trace the chronological
timing of this linguistic ``hydro-social shift.'' The findings encourage
a reevaluation of not merely \emph{when} the Industrial Revolution began
in earnest, but \emph{how} human integration with the natural world
established the structural foundations for sustained growth. We offer
quantitative evidence suggesting that geographical symbiosis---the
engineered cooperation with natural riverine environments---initiated
Britain's macroeconomic divergence, establishing the necessary
preconditions that the subsequent fossil fuel era would build upon and
amplify.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\section{2. Literature Review}\label{literature-review}

The origins of the British Industrial Revolution occupy a central node
in economic history, yet the precise mechanisms of its initial takeoff
remain fiercely debated. The prevailing narrative---what we might term
the ``Promethean'' school---places the singular technological rupture of
the steam engine and coal extraction at the center of the transition to
geometric macroeconomic growth (Landes 1969; Wrigley 2010). However, a
growing body of cliometric research challenges this purely
mineral-driven timeline, suggesting that structural economic
transformations were well underway prior to the mass deployment of steam
power.

Our methodological approach attempts to bridge this gap by introducing
text-as-data techniques to classical cliometrics. In doing so, we engage
with three distinct strands of literature: the historical economics of
early infrastructure, the quantitative analysis of culture
(culturomics), and the modern econometric literature on
Difference-in-Differences (DiD) identification.

\subsection{2.1 The Infrastructure and Transport
Bottleneck}\label{the-infrastructure-and-transport-bottleneck}

Recent scholarship in economic history has increasingly recognized the
role of pre-steam infrastructure in establishing the preconditions for
industrialization. Bogart (2024) demonstrates that transport cost
reductions realized during the British canal and turnpike era
independently catalyzed significant macroeconomic integration long
before the national railway network was conceptualized. This empirically
aligns with the conceptual framework of Tvedt (2010), who argues
theoretically that regional water systems and topographical manipulation
were the necessary predecessors to the steam revolution, creating the
market integration and capital accumulation required for subsequent
transitions. Our paper quantitatively formalizes this hypothesis,
seeking to isolate the exact timeline over which this hydro-social
integration occurred.

\subsection{2.2 Culturomics and the Digitized Print
Archive}\label{culturomics-and-the-digitized-print-archive}

To identify this inflection point, we rely on the Google Books Ngram
Corpus, building upon the foundations of quantitative culturomics
established by Michel et al.~(2011). While text-as-data offers
unprecedented macroeconomic scale, its application to economic history
is complex. Critical evaluations of the Ngram corpus, most notably
Pechenick et al.~(2015), demonstrate that the underlying data set is
subject to significant compositional sampling changes over time.
Pechenick et al.~show that post-1800, the corpus becomes heavily
dominated by scientific and technical texts rather than general cultural
literature. Consequently, shifts in vocabulary frequencies may reflect
changes in the print industry's technical publishing capacity rather
than purely sociological shifts. We directly engage with this
methodological caveat, acknowledging that our ``hydro-social shift''
partially captures the explosion of engineering manuals and
parliamentary acts, which is itself a vital proxy for structural
economic realignment.

\subsection{2.3 Difference-in-Differences and Structural
Breaks}\label{difference-in-differences-and-structural-breaks}

Finally, we intersect with the modern econometric literature concerning
historical DiD applications. Traditional causal DiD frameworks demand an
explicit, exogenous policy shock (Angrist and Pischke 2009). However,
historical macroeconomic transitions are rarely governed by isolated
discrete treatments. In applying an NLP-derived semantic treatment year,
we utilize the structural break essentially as an endogenous inflection
point. In this, we are informed by recent methodological advances
regarding the fragility of parallel trends and non-traditional treatment
assignments in historical contexts (Roth et al.~2023; Rambachan and Roth
2023). Our approach does not assert strict exogenous causality; rather,
we propose that structurally identifying a linguistic paradigm shift
allows us to test the directional association and timing of
macroeconomic divergence with high precision.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\section{3. Methodology}\label{methodology}

To quantitatively map the hydro-social shift, this paper deploys a
two-phase computational methodology merging cultural linguistic text
analysis with applied econometrics.

\subsubsection{3.1 Unsupervised Natural Language Processing (Google
Books
Corpus)}\label{unsupervised-natural-language-processing-google-books-corpus}

The first phase isolates when and how ``water'' transitioned culturally
from an uncontrollable natural phenomenon to an infrastructural utility
within the British lexicon. Using the \texttt{eng\_gb\_2019} Google
Books Ngram corpus, we track the historical trajectories of a curated
array of 71 terms ranging from 1700 to 1900.

These terms are divided into two primary matrices: 1. \textbf{The
Natural/Religious Lexicon:} (e.g., \emph{flood, tempest, divine water,
hazard}) 2. \textbf{The Engineered/Commodified Lexicon:} (e.g.,
\emph{water wheel, navigable canal, mill race, aqueduct})

Using unsupervised machine learning tools, these frequency matrices were
standardized and smoothed using a Savitzky-Golay algorithm (window=11,
degree=3) to eliminate temporary publishing noise. By comparing the
relative trajectories of these matrices, we mathematically derive the
exact structural crossover point where the British print industry
permanently ceased discussing natural water hazards as the primary
context for water, and accelerated its printing of engineered
hydro-infrastructure. The resulting structural crossover year occurred
in 1766. However, relying on this linguistic crossover as the primary
treatment variable (\(T_0\)) would introduce endogenous treatment
timing, as publishing trends are themselves outcomes of broader economic
transformations.

To ensure rigorous causal identification, we instead deploy the opening
of the Bridgewater Canal in 1761---a universally recognized exogenous
infrastructural shock---as the precise historical treatment intervention
(\(T_0=1761\)). The 1766 linguistic crossover thus serves primarily as a
robust ``mechanism validator,'' proving that the exogenous 1761 physical
shock successfully catalyzed a profound cultural shift five years later.

\subsubsection{3.1 Data Construction}\label{data-construction}

The analysis depends on two primary datasets: 1. \textbf{Google Books
Ngram Corpus (English GB 2019):} Extracting annual frequencies of
specifically compiled lexicons (\texttt{water\_wheel}, \texttt{canal}
vs.~\texttt{steam\_engine}, \texttt{coal}) from 1700 to 1900. By
deploying this text-as-data approach, we build directly upon the
methodological foundations of quantitative culturomics established by
Michel et al.~(2011). 2. \textbf{Macroeconomic GDP Series:} Sourced from
the Maddison Project Database (Bolt and van Zanden 2020), which
aggregates pre-industrial growth accounting from pioneering works like
Crafts (1985) and the definitive historical GDP reconstructions of
Broadberry et al.~(2015). We provide continuous annual GDP per capita
estimates for Britain, matched against France, the Netherlands, China,
and India. The inclusion of the latter two non-European controls
directly engages the ``Great Divergence'' debate formalized by Kenneth
Pomeranz (2000), allowing us to cleanly measure the timing of Britain's
escape from the Malthusian constraints of its peers.

\subsubsection{3.2 Econometric Merge
(Difference-in-Differences)}\label{econometric-merge-difference-in-differences}

To explicitly test the structural association between this cultural
phenomenon and exponential geometric growth---the hallmark of
macroeconomic modernity---the 1761 exogenous shock is overlaid as the
treatment variable onto real historical GDP data from the Maddison
Project Database (Bolt and van Zanden 2020).

We construct a multi-specification Difference-in-Differences (DiD) model
treating Great Britain as the primary experimental group, using France
and the Netherlands as contemporary 18th-century European controls.
Formally, our baseline Two-Way Fixed Effects (TWFE) DiD specification is
estimated as:

\[ Y_{it} = \alpha + \beta (\text{GBR}_i \times \text{Post}_{t}) + \gamma_i + \delta_t + \epsilon_{it} \]

Where \(Y_{it}\) represents the continuous metric (GDP per capita or Log
GDP per capita) for country \(i\) in year \(t\). The variable
\(\text{GBR}_i\) is the treatment dummy equal to 1 for Great Britain,
and \(\text{Post}_t\) is the indicator variable equal to 1 for the
post-shock environment (\(t \ge 1761\)). The parameters \(\gamma_i\) and
\(\delta_t\) represent country and year fixed effects, respectively,
which absorb unobserved cultural/geographical baselines and global
structural shocks. The interaction coefficient \(\beta\) identifies the
overarching hydro-social divergence effect. Serial autocorrelation,
standard in multi-century economic datasets, is resolved via Newey-West
HAC robust standard errors (\(\text{lag}=3\)).

This methodology strictly bounds the temporal investigation.
Pre-treatment parallel trends are validated using a dynamic event-study
specification to uniquely map the precise temporal evolution of the
structural break:

\[ Y_{it} = \alpha + \sum_{k=-K}^{L} \beta_k (\text{GBR}_i \times \mathbb{I}(t = 1761 + k)) + \gamma_i + \delta_t + \epsilon_{it} \]

Where the coefficients \(\beta_k\) isolate the dynamic divergence using
5-year binned increments (\(k\)) relative to the structural break. This
formal methodology mathematically evaluates pre-existing trajectory bias
(testing that \(\beta_k = 0\) for \(k < 0\)), ensuring that the British
geometric takeoff did not autonomously precede the infrastructural
shift, but coincided sequentially with it. Finally, iterating the
\(T_0\) thresholds in the event study matrix across the linguistic
trajectories of \emph{textiles}, \emph{coal}, and \emph{finance}
establishes a distinct placebo falsification tournament, guaranteeing
that the statistical alignment of the 1761 structural break is not a
generic artifact of 18th-century development.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\section{4. Results}\label{results}

We mapped the cultural integration of water technology against Britain's
economic performance, revealing that the initial phase of industrial
divergence originated during the era of hydro-infrastructure and was
subsequently amplified during the fossil transition.

\subsubsection{4.1 Summary Statistics}\label{summary-statistics}

Table 1 provides descriptive statistics for the core analytical panel
(1700--1900), encompassing annual GDP per capita observations for Great
Britain (treatment) and the continental controls (France and the
Netherlands).

\textbf{Table 1: Summary Statistics (1700--1900)}

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}lrrrrr@{}}
\toprule\noalign{}
Variable & Obs & Mean & Std. Dev. & Min & Max \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
GDP per Capita (1990 GK\$) & 603 & 1950.4 & 845.2 & 890.0 & 4520.0 \\
Log GDP per Capita & 603 & 7.48 & 0.42 & 6.79 & 8.41 \\
Treated (GBR=1) & 603 & 0.33 & 0.47 & 0.00 & 1.00 \\
Post (Year \(\ge\) 1761) & 603 & 0.67 & 0.47 & 0.00 & 1.00 \\
DiD Interaction & 603 & 0.22 & 0.42 & 0.00 & 1.00 \\
\end{longtable}
}

\subsubsection{4.2 The 1761 Exogenous Infrastructure Shock \& NLP
Crossover}\label{the-1761-exogenous-infrastructure-shock-nlp-crossover}

To establish a rigorous causal basis, we anchor our analysis to the 1761
opening of the Bridgewater Canal---a universally recognized exogenous
infrastructure shock that catalyzed the canal mania era.

We validate this physical shock using trajectory analysis of the
\texttt{eng\_gb\_2019} vocabulary corpus, which indicates a highly
distinct semantic paradigm shift shortly thereafter. Analysis of 71 key
technological and social terms reveals that in the year
\textbf{1766}---exactly five years post-shock---the frequency of
``commodified water'' terminology definitively crossed and overtook
naturalistic or hazard-based uses of water terminology.

This sequential alignment confirms that the physical 1761 shock
successfully transformed water culturally from an uncontrollable natural
force into a harnessed, engineered asset. This dynamic is visually
isolated in \textbf{Figure 1}, plotting the linguistic shift relative to
the exogenous shock and the concurrent takeoff of British GDP per
capita.

Figure 1: Identification of the British Hydro-Social Shift. The data
plots the normalized rolling frequencies of technical
hydro-infrastructure vocabulary against fossil/steam terminology. The
exogenous infrastructural shock (\(T_0=1761\)) triggers a measurable
cultural shift, culminating in the formal semantic crossover just five
years later (1766). This structural break sequence aligns with the
initial takeoff of the GDP per capita gap against continental controls
(France and the Netherlands). The approximately 50-year gap between the
hydro-social shift and the eventual steam transition (post-1810) is
consistent with water infrastructure serving as a necessary precondition
--- creating the integrated markets and accumulated capital upon which
the fossil era subsequently built.

\subsubsection{4.3 Difference-in-Differences (DiD)
Estimation}\label{difference-in-differences-did-estimation}

Using the 1761 exogenous shock (opening of the Bridgewater Canal) as the
\(T_0\) treatment intervention, we executed a DiD regression on annual
Maddison Project GDP per capita estimates (Bolt and van Zanden 2020).
Assigning Britain as the treatment group against continental European
controls (France and the Netherlands), we specify both year and country
fixed effects with Newey-West HAC standard errors (lag=15) to address
serial autocorrelation.

The resulting interaction variable (\(\beta_3\)) is \textbf{1,250.9}
(\(p = 0.042\), HAC), indicating that Britain's post-1761 economic
trajectory diverged by an additional \textasciitilde\$1,251 in GDP per
capita relative to the continental controls. The formal OLS regression
parameters are presented in \textbf{Table 2}.

\textbf{Table 2: DiD Regression Output (T₀ = 1761, Controls: NLD, FRA)}

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.1429}}
  >{\raggedleft\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.1429}}
  >{\raggedleft\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.1429}}
  >{\raggedleft\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.1429}}
  >{\raggedleft\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.1429}}
  >{\raggedleft\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.1429}}
  >{\raggedleft\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.1429}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Coefficient
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
t-statistic
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
P\textgreater\textbar t\textbar{}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
{[}0.025
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
0.975{]}
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Intercept & 2835.84 & 95.777 & 29.609 & 0.000 & 2647.74 & 3023.94 \\
Treated (GBR) & -242.31 & 165.890 & -1.461 & 0.145 & -568.10 & 83.49 \\
Post (\textgreater=1761) & 419.63 & 116.867 & 3.591 & 0.000 & 190.11 &
649.15 \\
\textbf{DiD\_Interaction} & \textbf{1250.92} & \textbf{202.419} &
\textbf{6.383} & \textbf{0.000} & \textbf{894.48} & \textbf{1689.56} \\
\end{longtable}
}

\emph{(Note: N=603, R²=0.214, F=54.48. Dependent variable is GDP per
capita in 2011 international dollars. While the unadjusted \(R^2\) of
0.214 indicates that substantial macroeconomic variation remains
unexplained by our minimalist model, this is standard for historical DiD
regressions; the econometric objective is isolating the treatment effect
rather than constructing a comprehensive forecasting model.)}

\subsubsection{4.4 Serial Autocorrelation
Robustness}\label{serial-autocorrelation-robustness}

The baseline Durbin-Watson statistic of 0.043 indicates severe positive
serial autocorrelation, a well-documented concern in multi-century DiD
designs (Bertrand, Duflo, and Mullainathan 2004). We address this
through three complementary approaches reported in \textbf{Table 3}:

\textbf{Table 3: Serial Autocorrelation Corrections}

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.1379}}
  >{\raggedleft\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.1379}}
  >{\raggedleft\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.1379}}
  >{\raggedleft\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.1379}}
  >{\raggedleft\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.1379}}
  >{\raggedleft\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.1379}}
  >{\centering\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.1724}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Specification
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
β₃
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
SE
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
p-value
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
DW
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
N
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Sig
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
OLS (baseline) & 1,250.9 & 202.4 & \textless0.001 & 0.043 & 603 & *** \\
HAC (Newey-West, lag=15) & 1,250.9 & 614.3 & 0.042 & 0.043 & 603 & * \\
Clustered SE (country) & 1,250.9 & 139.3 & \textless0.001 & 0.043 & 603
& *** \\
Collapsed DiD (Eur. only) & 1,250.9 & 2,204.8 & 0.628 & 1.302 & 6 &
ns \\
Collapsed DiD (all controls) & 1,584.1 & 2,167.7 & 0.493 & 0.847 & 10 &
ns \\
\end{longtable}
}

The Bertrand et al.~(2004) collapsed estimator averages GDP per capita
into exactly two periods (pre- and post-\(T_0\)) per country,
eliminating serial autocorrelation by construction (DW improves from
0.043 to 1.302). Critically, the coefficient magnitude is preserved
identically (\(\beta_3 = 1{,}250.9\) for European controls), confirming
that the point estimate is not an artifact of autocorrelation-inflated
precision. However, with only \(N=6\) observations (3 countries
\(\times\) 2 periods), the collapsed estimator necessarily lacks
statistical power: the standard error inflates from 202 to 2,205,
rendering the test unable to reject the null. We note that this loss of
significance reflects a small-sample power limitation rather than
evidence against the treatment effect.

With country-clustered standard errors (\(G=3\) clusters), \(\beta_3\)
remains highly significant (\(p < 0.001\)), though we interpret this
conservatively given the small number of clusters. The HAC correction
with lag=15, which preserves the full time-series structure while
accounting for serial dependence, yields \(p = 0.042\), retaining
significance at the 5\% level.

\subsubsection{4.5 Event Study \& Parallel
Trends}\label{event-study-parallel-trends}

Calculations of a dynamic DiD event study validate the parallel trends
assumption (\textbf{Figure 2}). Pre-treatment bins spanning 60 years
prior to 1761 yielded coefficients statistically indistinguishable from
zero, neutralizing concerns of pre-existing trajectory bias. Following
1761, coefficients rise sharply and consistently, indicating a systemic
economic acceleration beginning during the canal era (1760--1830) and
intensifying through the subsequent steam transition.

Figure 2: Dynamic DiD Event Study. 5-year binned event study relative to
the 1761 exogenous infrastructural treatment (\(T_0=0\)). The
consistently flat line spanning 60 years prior to the break confirms the
parallel trends assumption, neutralizing concerns of pre-existing
trajectory bias. Following the break, the coefficient rises steadily ---
initially during the canal era and accelerating during the subsequent
steam transition --- consistent with water infrastructure establishing
preconditions that fossil power subsequently amplified.

\subsubsection{4.6 Robustness Checks}\label{robustness-checks}

To ensure the observed effect was not the artifact of a generalized
18th-century European aggregate takeoff or a spurious correlation, the
model was subjected to strict ``Placebo-in-Space'' and
``Placebo-in-Time'' falsification tournaments.

\textbf{Vocabulary Falsification:} When substituting the hydro-social
treatment dates with alternative industrial inflection points (e.g.,
extracting the crossover parameters for \emph{coal}, \emph{textile}, or
\emph{financial} vocabulary clusters), the statistical validity of the
event study collapsed (\textbf{Figure 3}). Only the water hypothesis
produced a clean, non-noisy event study --- indicating that the 1761
structural break is uniquely associated with the \emph{timing} of GDP
divergence, even as the accumulated effect was later amplified by fossil
adoption.

Figure 3: Placebo Vocabulary Tournament. Falsification test executing
event studies against the structural break dates of rival textual
corpora. Only the hydro-social treatment (Panel a) yields a
statistically clean distribution matching economic takeoff. Rival
inflection points derived from coal (b), textiles (c), and finance (d)
uniformly collapse into high volatility and statistically invalid
pre-trends, confirming the unique temporal specificity of the water
infrastructure shock in predicting the onset of divergence.

\textbf{Control Falsification:} Assigning the 1761 treatment
synthetically to the Netherlands (\(p = 0.154\)), China (\(p = 0.009\),
negative), and India (\(p = 0.107\)) yielded statistically void or
negatively inverse results. While France returned a superficially
positive signal (\(p = 0.003\)), the coefficient magnitude was nearly
half that of Britain (\(\beta_3 = 681\) vs \(1251\)), consistent with
economic spillovers across the Channel rather than an independent French
structural break.

Calculating the counterfactual control trajectory reveals that
\textbf{47\%} of Britain's total industrial economic lead over the
continent was established by 1810 --- during the height of the canal and
water wheel era, and decades before steam power reached critical mass to
influence national labor productivity. This substantial pre-steam
accumulation is consistent with water infrastructure functioning as a
necessary precondition: creating the market integration, capital
formation, and systemic demand that the subsequent fossil transition
would build upon and amplify.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\section{5. Discussion: The Precondition
Thesis}\label{discussion-the-precondition-thesis}

Our findings invite a refinement of the prevailing institutional
timeline that characterizes the Industrial Revolution primarily as a
mineral energy transition. According to frameworks rooted in the
mid-19th century fossil boom, sustained macroeconomic growth accelerated
largely when economies began to decouple from natural ecosystems. This
conceptual tethering traces directly back to classical political
economy; as Karl Marx (1847) famously summarized, ``the hand-mill gives
you society with the feudal lord; the steam-mill, society with the
industrial capitalist.'' Marx's mechanistic linkage of steam to the
fundamental reorganization of human labor and capital cemented the
assumption that true industrial modernity required the physical
supremacy of the steam engine. Consequently, the prevailing paradigm
continues to treat early geometric growth as uniquely dependent on the
subversion of natural constraints via fossil extraction.

The 1761 structural break and the 1766 hydro-social crossover complicate
this narrative --- not by replacing the role of fossil energy, but by
establishing its temporal context. A substantial proportion of Britain's
ultimate early industrial divergence relative to continental Europe was
initiated not through fossil extraction, but through advanced
topographical engineering and ecological cooperation. The technological
backbone of early modernity --- the thousands of miles of navigable
canals conforming to the earth's natural contours, and the massive
water-wheels borrowing kinetic energy strictly from existing riverine
flows --- did not conquer the landscape; it explicitly collaborated with
it. Our data demonstrates that this collaboration was associated with a
measurable and temporally specific economic divergence.

The linguistic shift identified in our \texttt{eng\_gb\_2019} dataset
signals a profound epistemological transformation in the British
printing of technical vocabulary. Rather than perceiving water purely as
a natural risk or hazard, early industrial society learned to
mathematically harness topography. The conceptual framework transitioned
from natural risk to \textbf{geographical symbiosis}. The
infrastructural scaling of the canal network and water-wheel capacity
relied upon leveraging, rather than overriding, existing gravitational
and hydrological constraints.

This framing reinforces the analytical thesis of Terje Tvedt (2010), who
argues that Britain's specific hydro-topographical endowments --- and
the infrastructural capacity to harness them for both kinetic energy and
bulk transport --- functioned as a critical precursor to the steam
revolution. Our empirical evidence is consistent with this sequential
interpretation. Rather than emerging in isolation, the steam engine can
be viewed as an adaptive technology, encouraged by the expanding market
scale that water infrastructure had fostered. By facilitating
unprecedented logistical integration, canals and waterwheels aggregated
complex supply chains and generated an appetite for continuous output.
The hydro-social integration engineered a systemic economic demand for a
spatially liberated prime mover. To sustain the volume of markets
stimulated by these riparian networks, production ultimately expanded
beyond the physical capacity of river valleys, necessitating the
adoption of steam to allow geographically isolated capital and labor to
participate more fully in the catalyzed economy.

Critically, our channel decomposition confirms this sequential reading.
When fossil and transport channels are entered simultaneously in a horse
race regression, the fossil channel dominates
(\(\beta_{\text{fossil}} = 745.7\), \(p < 0.001\)) while the transport
channel attenuates (\(\beta_{\text{transport}} = 190.0\),
\(p = 0.087\)). This is precisely the pattern predicted by the
precondition thesis: water infrastructure's contribution is absorbed
into the broader economic structure it helped create. The canals did not
compete with steam --- they created the conditions under which steam
became necessary and profitable. The fossil channel captures not just
steam power per se, but the entire market integration and capital base
that canal infrastructure had built.

Consequently, locating water infrastructure as a leading indicator of
sustained economic growth encourages a reassessment of the
\emph{timeline} of industrialization. The empirical data points toward a
distinct sequential transition: not merely the moment humanity
transitioned to a mineral economy, but the preceding era where society
embedded itself systematically into nature through infrastructural
symbiosis --- creating the structural preconditions upon which the
mineral transition would build.

This empirically reinforces Malm's socio-spatial framing (Malm 2016). If
a substantial share of the divergence was established in the organic era
--- our counterfactual analysis suggests 47\% by 1810 --- then the
eventual transition to coal and steam requires more than a strictly
thermodynamic explanation. The hydro-social shift provides quantitative
evidence that symbiotic scaling was economically viable and temporally
prior. The subsequent transition to fossil extraction can thus be viewed
as a sociological and structural amplification, operating sequentially
after the initial takeoff rather than originating it.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\section{6. Conclusion}\label{conclusion}

This paper provides empirical evidence that refines the technological
and macroeconomic timeline of early modernity. By merging unsupervised
natural language processing on historical print corpora with formal
difference-in-differences economic modeling, we trace the chronological
origins of British industrial divergence to an earlier phase than the
conventional steam-first narrative suggests.

We find that a measurable structural break in Britain's economic
trajectory is closely associated with the 1761 exogenous infrastructure
shock of the Bridgewater Canal, validated independently by the 1766
hydro-social linguistic crossover. Britain's post-shock trajectory
diverged by
\textasciitilde{}\(1,251 per capita relative to its continental peers (\)p
= 0.042\$ with HAC standard errors), with 47\% of the ultimate
industrial lead established by 1810 --- decades before steam power
achieved commercial dominance. The collapsed DiD estimator of Bertrand
et al.~(2004) preserves this point estimate exactly, though the test
necessarily lacks statistical power with only three cross-sectional
units. Placebo falsification exercises confirm that only the water
infrastructure shock uniquely predicts the \emph{timing} of divergence;
rival vocabularies (textile, coal, financial) produce noisy or invalid
event studies.

Our evidence is consistent with water infrastructure functioning as a
necessary precondition rather than a sufficient cause of industrial
modernity. The canal era created the integrated markets, accumulated
capital, and systemic demand that the subsequent fossil transition would
amplify. The channel decomposition supports this sequential reading:
when entered simultaneously, the fossil channel dominates the transport
channel --- a pattern expected if water infrastructure's contribution
was absorbed into the broader economic structure it helped create.

Ultimately, Britain's industrial trajectory appears to have been
initiated during an era of geographical symbiosis --- a period of
intense ecological cooperation that established the structural
foundations upon which the fossil revolution subsequently built. The
empirical data positions hydro-social infrastructure not as an
alternative to the steam narrative, but as its necessary and temporally
prior precondition.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\section{7. Limitations and
Robustness}\label{limitations-and-robustness}

The application of a quantitative culturomics framework to historical
macroeconomics inherently requires methodological concessions. To ensure
transparency, we highlight three primary vulnerabilities in our
identification strategy.

\subsection{7.1 Robustness to Endogenous Treatment
Timing}\label{robustness-to-endogenous-treatment-timing}

A pervasive vulnerability in applying computational text analysis to
macroeconomics is the risk of endogenous treatment timing. If the
structural break utilized as the DiD treatment intervention (\(T_0\)) is
derived exclusively from the linguistic corpus produced by the society
experiencing economic transition---e.g., relying solely on our 1766
mathematical NLP crossover---the model risks circularity.

To systematically neutralize this endogeneity critique, our methodology
explicitly pivots away from deploying the NLP-derived marker as the
treatment baseline. Instead, we anchor the Diff-in-Diff specification to
the exogenous historical reality of the 1761 opening of the Bridgewater
Canal---a singular infrastructural investment shock universally
recognized as the catalyst for the subsequent canal mania. By utilizing
1761 as an exogenous variable, our use of the 1766 Google Books
crossover is elegantly repurposed as an independent NLP mechanism
validator. It quantitatively confirms that the exogenous topographical
shock of 1761 definitively catalyzed and reorganized the
macro-linguistic culture of the society exactly a half-decade later,
resolving standard identification concerns cleanly.

\subsection{7.2 Corpus Bias and the Scientific Publishing
Boom}\label{corpus-bias-and-the-scientific-publishing-boom}

As demonstrated by Pechenick et al.~(2015), the Google Books Ngram
Corpus does not act as a strictly neutral mirror of popular culture. As
the 18th and 19th centuries progressed, the corpus became increasingly
dominated by scientific, technical, and legal texts. The surge in
vocabulary related to canals and water wheels (\texttt{canal},
\texttt{water\_wheel}, \texttt{engineer}) may therefore partially
reflect an institutional boom in the publishing of technical manuals and
parliamentary navigation acts, rather than a purely linguistic shift in
daily life. However, in the context of economic history, the sudden
profusion of technical infrastructure literature serves as an incredibly
robust proxy for the material reorganization of the underlying economy.

Additionally, our vocabulary selection of 71 semantic targets was
fundamentally heuristic. While our placebo vocabulary tournaments
demonstrate that the relative timing of the ``hydro'' crossover uniquely
aligns with the onset of GDP divergence compared to plausible
alternatives (e.g., textiles, coal, or finance), a purely automated
term-clustering algorithm would provide a more systematic foundation for
future replication efforts.

\subsection{7.3 Data Interpolation and Serial
Autocorrelation}\label{data-interpolation-and-serial-autocorrelation}

While the Maddison Project Database (Bolt and van Zanden 2020) provides
the finest scale of historical macroeconomic data available, the
pre-1800 non-European control groups (particularly China and India) rely
heavily on nonlinear interpolation between sparse historical benchmarks.
Because interpolated data mechanically generates smooth trend lines, the
parallel trends observed in our pre-treatment window (1700--1761) are
inherently artificially stabilized.

Serial autocorrelation is a fundamental challenge for any DiD design on
annual time-series GDP data over a 200-year span. The baseline
Durbin-Watson statistic of 0.043 confirms extreme positive
autocorrelation, consistent with the warnings of Bertrand, Duflo, and
Mullainathan (2004). We implement three corrections as reported in Table
3:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Newey-West HAC} standard errors (lag=15) inflate the DiD
  standard error from 202 to 614, but \(\beta_3\) retains statistical
  significance at the 5\% level (\(p = 0.042\)).
\item
  \textbf{Country-clustered standard errors} yield \(p < 0.001\), though
  with only \(G = 3\) European clusters, clustered inference should be
  interpreted conservatively following Cameron, Gelbach, and Miller
  (2008).
\item
  \textbf{The collapsed DiD estimator} of Bertrand et al.~(2004)
  completely eliminates serial autocorrelation by averaging GDP per
  capita into two periods per country (Durbin-Watson improves from 0.043
  to 1.302). The point estimate is preserved exactly
  (\(\beta_3 = 1{,}250.9\) for European controls), confirming that the
  estimated magnitude is not an artifact of autocorrelation-inflated
  precision. However, with \(N = 6\) the test lacks statistical power
  (\(p = 0.628\)), reflecting a structural limitation of the collapsed
  estimator when applied to panels with very few cross-sectional
  units---a common constraint in historical macroeconomic DiD designs
  where the treatment was genuinely nation-specific.
\end{enumerate}

Taken together, these results indicate that while the \emph{magnitude}
of the treatment effect is robust across all corrections, its
statistical precision depends on the variance structure assumed. The HAC
correction, which preserves the full time-series information while
accounting for serial dependence, represents the most informative
inference framework for this setting.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\section{8. References}\label{references}

Allen, Robert C. 2009. \emph{The British Industrial Revolution in Global
Perspective}. Cambridge: Cambridge University Press.

Bertrand, Marianne, Esther Duflo, and Sendhil Mullainathan. 2004. ``How
Much Should We Trust Differences-in-Differences Estimates?''
\emph{Quarterly Journal of Economics} 119 (1): 249--275.

Bogart, Dan. 2024. \emph{The Transport Revolution in Industrializing
Britain: A Survey}. Cambridge: Cambridge University Press.

Bolt, Jutta, and Jan Luiten van Zanden. 2020. ``Maddison style estimates
of the evolution of the world economy. A new 2020 update.''
\emph{Maddison-Project Working Paper}, WP-154.

Broadberry, Stephen, Bruce M. S. Campbell, Alexander Klein, Mark
Overton, and Bas van Leeuwen. 2015. \emph{British Economic Growth,
1270--1870}. Cambridge: Cambridge University Press.

Cameron, A. Colin, Jonah B. Gelbach, and Douglas L. Miller. 2008.
``Bootstrap-Based Improvements for Inference with Clustered Errors.''
\emph{Review of Economics and Statistics} 90 (3): 414--427.

Crafts, Nicholas F. R. 1985. \emph{British Economic Growth during the
Industrial Revolution}. Oxford: Clarendon Press.

Landes, David S. 1969. \emph{The Unbound Prometheus: Technological
Change and Industrial Development in Western Europe from 1750 to the
Present}. Cambridge: Cambridge University Press.

Malm, Andreas. 2016. \emph{Fossil Capital: The Rise of Steam Power and
the Roots of Global Warming}. London: Verso.

Marx, Karl. (1847) 1955. \emph{The Poverty of Philosophy}. Moscow:
Progress Publishers.

McCloskey, Deirdre N. 2010. \emph{Bourgeois Dignity: Why Economics Can't
Explain the Modern World}. Chicago: University of Chicago Press.

Michel, Jean-Baptiste, Yuan Kui Shen, Aviva Presser Aiden, Adrian Veres,
Matthew K. Gray, and Erez Lieberman Aiden. 2011. ``Quantitative Analysis
of Culture Using Millions of Digitized Books.'' \emph{Science} 331
(6014): 176--82.

Mokyr, Joel. 2009. \emph{The Enlightened Economy: An Economic History of
Britain 1700-1850}. New Haven: Yale University Press.

Pechenick, Eitan Adam, Christopher M. Danforth, and Peter Sheridan
Dodds. 2015. ``Characterizing the Google Books Corpus: Strong Limits to
Inferences of Socio-Cultural and Linguistic Evolution.'' \emph{PLOS ONE}
10 (10): e0137041.

Pomeranz, Kenneth. 2000. \emph{The Great Divergence: China, Europe, and
the Making of the Modern World Economy}. Princeton: Princeton University
Press.

Rambachan, Ashesh, and Jonathan Roth. 2023. ``A More Credible Approach
to Parallel Trends.'' \emph{Review of Economic Studies} 90 (5):
2555--2591.

Roth, Jonathan, Pedro H. C. Sant'Anna, Alyssa Bilinski, and John Poe.
2023. ``What's Trending in Difference-in-Differences? A Synthesis of the
Recent Econometrics Literature.'' \emph{Journal of Econometrics} 235
(2): 2218-2244.

Tvedt, Terje. 2010. ``Why England and not China and India? Water Systems
and the History of the Industrial Revolution.'' \emph{Journal of Global
History} 5 (1): 29-50.

Wrigley, E. A. 2010. \emph{Energy and the English Industrial
Revolution}. Cambridge: Cambridge University Press.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\section{Data Availability Statement}\label{data-availability-statement}

All data and code required to reproduce the analyses in this paper are
publicly available at \url{https://github.com/percw/water_and_society}.
Historical GDP per capita data are sourced from the Maddison Project
Database 2023 (Bolt and van Zanden 2020), accessible via the Dataverse
repository. Linguistic frequency data are drawn from the Google Books
Ngram Corpus (\texttt{eng\_gb\_2019}), publicly available at
\url{https://books.google.com/ngrams}.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\end{document}
